The effect of clumpy outflows and disc anisotropy on quasar unification scenarios

### 2015-07-15 paper.tex
a bunch of text in results, may clash with laptop

%% Lines and ions
\def\civ{C~\textsc{iv}}
\def\nv{N~\textsc{v}}
\def\heii{He~\textsc{ii}}
\def\ovi{O~\textsc{vi}}
\def\la{Ly$\alpha$}
\def\ha{H$\alpha$}

@@

Introduction focussing on key points

\begin{itemize}
\item standard wind + BALQSO introduction
\item focus on unification and that we are testing it
\item some discussion of scales, referencing e.g. reverb maps, variability, microlensing, Arav 
\item clumping background: stellar winds, clumping in AGN winds, variability
\end{itemize}
%%\item

\clearpage
Quasars and luminous active galactic nuclei (AGN) 
exhibit spectral energy distributions (SEDs) that
typically exhibit a series of strong emission lines (e.g. \la, \civ, \nv) 
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is normally attributed to 
emission from a geometrically thin, optically thick accretion disc 
surrounding the central black hole (REF), similar
to that described by \cite{shakurasunyaev1973}.

@@

We use the same atomic data as described by LK02 and since
updated by H13 and M15, with the addition of direct ionization data from Dere (????). 
Photoionization cross-sections are from Topbase (REF) and 
\cite{vfky}.

@@

\section{A Clumpy Biconical Disk Wind Model for QSOs}

Our kinematic prescription for a biconical disc wind model
follows \cite{SV93}, and is described further by
LK02, H13 and M15.

@@

Describe the photon sources in the model.

The accretion disc in our model is geometrically thin, but optically thick
and this adopt a standard multi-temperature blackbody
using a \cite{shakurasunyaev1973} temperature profile. 
The emergent SED is thus determined by the specified accretion rate ($\dot{m}$)
and central BH mass ($M_{BH}$).
The inner radius of the disc extends to the innermost stable circular orbit
(ISCO) of the BH. 
%% OUTER RADIUS? 
We assume a Schwarzchild BH with an ISCO at $6~R_G$.
The X-ray source in our model is treated as an isotropic sphere at the ISCO,
which emits photons according to a power law,

% \begin{equation}

@@

\clearpage

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/wind.eps}
\caption
{
Physical properties of the outflow.
}
\label{fig:uvspec}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/uvspec.eps}
\caption
{
Synthetic spectra at four viewing angles in the clumpy model. Plot would look different probably,
and may show composites, but this would be the main synthetic spectrum plot.
}
\label{fig:uvspec}
\end{figure*}

@@

Show that clumping stops over-ionization.

{\bf Figure 2: Physical conditions of the wind, showing e.g. CIV fraction.}

@@

{\bf Figures 3 and 4: UV and optical spectra.}

\subsection{X-ray Properties}

discuss the figure showing X-ray properties briefly. 
Present an X-ray spectrum? compare to observations e.g. Giustini?

{\bf Figures 5 and 6: $L_{2kev}$ v $L_{2500}$ plot and X-ray spectrum compared to Grupe and Nousek (broadband SED?)}

@@

{\bf Figure 7: line profiles of CIV, Mg II and Al III as a function of inclination.}

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/c4_with_i.eps}
\caption
{
Carbon IV, Al III and Mg II line profiles with inclination for wind angles.
Showing how the line profiles change
and how LoBALs kick in at particularly high inclinations.
}
\label{fig:xray}
\end{figure*}

@@

\begin{enumerate}
\item We have introduced a simple, first-order treatment 
of clumping in our model, and found that it can now maintain
the required ionization state while agreeing well with the X-ray
properties of AGN/QSOs.
\smallskip
\item We find that clumping also causes a significant 
increase in the strength of the  emission
lines produced by the model. This is true both
of collisionally excited resonance lines (such as \civ, \nv)
and recombination lines (such as \la, \ha\ and the Balmer series).
\smallskip
\item The line EWs in our models are not comparable to those in Quasar composite
spectra. This is due to a fundamental constraint discussed in section ?. If the BLR
emits fairly isotropically then for a foreshortened, limb-darkened classical thin accretion disc
it is not possible to achieve line ratios at low inclinations that are comparable to
those at high inclinations. This is a robust conclusion which 
is independent of the assumed BLR geometry. 
\smallskip
\item We have examined the effect of GR on our disc SED, using the disc atmosphere
and GR ray-tracing code \agn. While including GR effects
does cause the disc SED to become significantly more isotropic,
the effect is not large enough to produce uniform line to continuum ratios
with viewing angle. We discuss other solutions to this problem in section ?; 
It is possible that a combination of GR and reprocessing by the wind could provide a 
solution, and a number of complicated selection effects may be at work
in the building of the quasar composites.
\end{enumerate}

@@

>>>>

%% Lines and ions
\def\civ{C~\textsc{iv}}
\def\nv{N~\textsc{v}}
\def\hei{He~\textsc{i}}
\def\heii{He~\textsc{ii}}
\def\mg{Mg~\textsc{ii}}
\def\al{Al~\textsc{iii}}
\def\heii{He~\textsc{ii}}
\def\ovi{O~\textsc{vi}}
\def\la{Ly~$\alpha$}
\def\ha{H~$\alpha$}
\def\hb{H~$\beta$}

@@

% Introduction focussing on key points
% \begin{itemize}
% \item standard wind + BALQSO introduction
% \item focus on unification and that we are testing it
% \item some discussion of scales, referencing e.g. reverb maps, variability, microlensing, Arav 
% \item clumping background: stellar winds, clumping in AGN winds, variability
% \end{itemize}
% %%\item 
% \clearpage

The spectral energy distributions (SEDs) of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines (e.g. \la, \civ, \nv) 
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is normally attributed to emission from a geometrically thin, 
optically thick accretion disc surrounding the central black hole (REF), similar
to that described by \cite{shakurasunyaev1973}.

@@

We use the same atomic data as described by LK02 and since
updated by H13 and M15, with the addition of direct ionization data from Dere (YEAR). 
Photoionization cross-sections are from Topbase (REF) and  \cite{vfky}.

@@

\section{A Clumpy Biconical Disk Wind Model for Quasars}

Our kinematic prescription for a biconical disc wind model
follows \cite{SV93}, and is described further by
LK02, H13 and M15. A schematic is shown in figure~\ref{fig:cartoon},
with key aspects marked. The general biconical
geometry is similar to that invoked by \cite{MCGV95} and 
\cite{elvis2000} in order to explain the phenomenonology
of quasars and BALQSOs.

@@

%Describe the photon sources in the model.

The accretion disc in our model is geometrically thin, but optically thick
and this adopt a standard multi-temperature blackbody
using a \cite{shakurasunyaev1973} temperature profile. 
The emergent SED is thus determined by the specified accretion rate ($\dot{m}$)
and central BH mass ($M_{BH}$).
The inner radius of the disc extends to the innermost 
stable circular orbit (ISCO) of the BH. 
We assume a Schwarzchild BH with an ISCO at $6~R_G$.
The X-ray source in our model is treated as an isotropic sphere at the ISCO,
which emits photons according to a power law with index $\alpha_X$.
The normalisation of this power law is such that it produces the specified 
2-10kev luminosity.
In addition to the disc and X-ray source, 
the wind is able to reprocess radiation. However, new 
photon packets are not produced in the wind (as in LK02). 
Instead, this reprocessing is dealt with by enforcing strict
radiative equilibrium via an indivisible energy packet
constraint (see Lucy 2002, M15).
% \begin{equation}

@@

@@

Here we describe the results from our model.

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/wind.eps}
\caption
{
Physical properties of the outflow, shown by the coloured contours.
The solid black line marks a sphere at $1000~R_G$.
The dotted lines show the $75^\circ$ and $80^\circ$ sightlines 
to the centre of the system, and illustrate that different sightlines
intersect material of different ionization states.
}
\label{fig:uvspec}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/sightlines.eps}
\caption
{
The classes of sightlines and the corresponding possible 
system types in the framework of our model.
[FIGURE NEEDS IMPROVING]
}
\label{fig:lobal}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{figures/uvspec.eps}
\caption
{
Synthetic spectra at four viewing angles in the clumpy model. Plot would look different probably,
and may show composites, but this would be the main synthetic spectrum plot.
}
\label{fig:uvspec}
\end{figure*}

@@

The wind rises slowly from the disc at first, with clumped densities
of $n_H \sim 10^{11}~\rm{cm^{-3}}$ close to the disc plane.
The flow then accelerates over a scale length of $R_V=10^{19}~\rm{cm}$
up to a terminal velocity of around $3$ times the escape velocity 
($\sim10,000~\rm{km~s^{-1}}$). This gradual acceleration means that
the wind exhibits a stratified ionization structure, with low ionization material
in the base of the wind giving way to highly ionized plasma further out.
By clumping the wind, we are able to produce the range of ionization states observed
in Quasars and BALQSOs, while adopting a realistic $2-10$ keV X-ray luminosity
of $L_{X}=10^{43}~\rm{ergs~s^{-1}}$. Without clumping, this wind would be over-ionized 
to the extent that opacities in e.g., \civ\ would be entirely negligible (see H13).

One commonly used measure of the ionization state is the ionization parameter, $U$, given by
\begin{equation}
U = \frac{4\pi}{n_H c}\int_{13.6{\rm{eV}}}^{\infty}\frac{{J_{\nu}d\nu}}{h\nu}.
\end{equation}
\noindent where $\rm{n_H}$ is the local number density of H, and $\nu$ denotes photon 
frequency. The ionization parameter represents the ratio of the number density of 
ionizing photons to the local H density. It is however, a poor measure of the 
ionization state of the resonance species such as \civ\ as it encodes no information
about the shape of the SED. In this case the X-ray photons 
are dominant in the photoionization of the UV resonance line ions. 
This explains why a factor of 100 increase in X-ray luminosity requires
a clumping factor of 0.01, even though the value of U changes by roughly ?? compared to 
H13.

Clumping also causes the total line luminosity to increase dramatically,
as recombination and collisional excitation are both proportional to
$n_e^2$. This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the X-rays heating the plasma.
The radii of these emitting regions is important,
and can be compared to observations. Our \civ\ line emission region 
is located at about $1000~R_G$ ($\sim10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civ\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

\subsection{X-ray Properties and Broadband SEDs}

The bottom panel of figure~\ref{fig:xray} shows the emergent $L_{2keV}$ 
monochromatic luminosity ($L_\nu$) and $\alpha_{OX}$ plotted against 
$L_\nu$ at $2500$\AA\ for a number of different viewing angles in our model.
$\alpha_{OX}$ is a spectral index from near UV to X-rays defined by

\begin{equation}
\alpha_{OX}=0.3838\log\left(\frac{L_{\nu}(2~keV)}{L_{\nu}(2500~\mbox{\scriptsize{\AA}})}\right).
\end{equation}

The properties are calculated from the synthetic spectra and thus include
the effects of wind reprocessing and attenuation. In addition to model outputs,
we also show the BALQSO sample of Saez et al. (2012) and luminous AGN and quasar
samples from Steffen et al. (2006). For low inclination, `quasar-like' viewing angles,
we now show excellent agreement with AGN samples. The trend with inclination
in our models is caused by a combination of disc foreshortening/limb-darkening 
(resulting in a lower $L_{2500}$ for higher inclinations) and the fact that the disk 
is opaque and thus the X-ray source subtends a smaller solid angle at high inclinations
(resulting in a lower $L_{2keV}$ for higher inclinations).

Although the low inclination, `BALQSO-like' viewing angles show moderate agreement with the data,
it appears that our models tend to underpredict the emergent X-ray luminosity at 
although we are limited by poor statistics. It appears that our models tend
It is possible that this is due to our wind being overly optically thick to 
electron scattering at angles which look through the wind. It could also be that the
isotropic X-ray source assumption is incorrect and BALQSOs are {\em intrinsically} 
X-ray weak, as suggested by REFS. Nevertheless, our input X-ray spectrum
now reproduces the X-ray properties of a luminous quasar.

We can also examine the emergent X-ray spectra. 
Figure 6 shows the broadband SED from the model, compared to a number of BALQSO spectra
from Grupe \& Nousek (2015) - note that lower mass and super-Eddington sources
are excluded. Our code is not yet optimized for producing X-ray spectra, as we do not yet 
include all lines and levels of e.g. Fe or macro-atom treatments of 
higher order ions (cf. Sim et al. 2008, 2010). 
However, the bound-free opacity sources are complete, and comparisons of the shape
of the heavily absorbed spectra can be useful.

% discuss the figure showing X-ray properties briefly. 
% Present an X-ray spectrum? compare to observations e.g. Giustini?

% {\bf Figures 5 and 6: $L_{2kev}$ v $L_{2500}$ plot and X-ray spectrum compared to Grupe and Nousek (broadband SED?)}

@@

At certain sightlines, our model now produces blue-shifted BALs in \al\ and \mg --
the absorption lines seen in LoBALQSOs. Line profiles in velocity space 
for \civ, \al\ and \mg, are shown in figure~\ref{fig:lobal} for a range
of BALQSO viewing angles. We confirm the behaviour expected from 
a unification model such as Elvis (2000), in which ionization stratification
explains the incidence of LoBALQSOs. In this framework, the lower
ionization material is intercepted by a smaller family of sightlines.
This is also the explanation here, as shown by figure~\ref{fig:wind}.

These line profiles also indicate a potential problem with our model. 
\cite{odowd2015} find that the LoBAL profiles in H1413+117 have a 
higher velocity onset than the higher ionization BALs, suggesting that the
wind becomes less ionized as radius increases. 
We find the opposite trend in velocity onsets. In addition, examination of 
figure~\ref{fig:wind} shows that ionization parameter decreases with radius, 
because the decrease in density wins over geometric dilution of the radiation 
field. While the \cite{odowd2015} hypothesis comes from one object,
it is clear that {\em at least some} BALQSOs must become less ionized at 
higher velocities, contrary to what one expects from our modelling.

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/c4_with_i.eps}
\caption
{
Carbon IV, Al III and Mg II line profiles with inclination for wind angles.
Showing how the line profiles change
and how LoBALs kick in at particularly high inclinations.
}
\label{fig:lobal}
\end{figure*}

@@

\begin{enumerate}
\item We have introduced a first-order treatment 
of clumping in our model, and found that it can now maintain
the required ionization state while agreeing well with the X-ray
properties of AGN/QSOs.
\smallskip
\item We find that clumping also causes a significant 
increase in the strength of the  emission
lines produced by the model. This is true both
of collisionally excited resonance lines (such as \civ, \nv)
and recombination lines (such as \la, \ha\ and the Balmer series).
\smallskip
\item The line EWs in our models are not comparable to those in Quasar composite
spectra. This is due to a fundamental constraint discussed in section ?. If the BLR
emits fairly isotropically then for a foreshortened, limb-darkened classical thin accretion disc
it is not possible to achieve line ratios at low inclinations that are comparable to
those at high inclinations. This is a robust conclusion which 
is independent of the assumed BLR geometry. 
\smallskip
\item We have examined the effect of GR on our disc SED, using the disc atmosphere
and GR ray-tracing code \agn. While including GR effects
does cause the disc SED to become significantly more isotropic,
the effect is not large enough to produce uniform line to continuum ratios
with viewing angle. We discuss other solutions to this problem in section ?; 
It is possible that a combination of GR and reprocessing by the wind could provide a 
solution, and a number of complicated selection effects may be at work
in the building of the quasar composites.
\end{enumerate}

@@

We would like to thank Omer Blaes, Ivan Hubeny, Shane Davis, 
Michael Crenshaw, Nahum Arav, Daniel Proga,
